\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-08-04 analysis/paper/paper.tex
fixed plot breaks and litrev. names

Several packages implement penalized estimation for the Cox model:
\pkg{glmnet} \citeyearpar{regpathcox}, \pkg{glmpath}
\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
\pkg{RiskRegression} \citeyearpar{gerds_blanche}. Moreover, some
packages also include penalized estimation in the context of Cox models
with time-varying coefficients: elastic-net penalization with
\pkg{CoxRidge} \citeyearpar{perperoglou} and \pkg{rstpm2}
\citeyearpar{clements_liu}, while \pkg{survival}
\citeyearpar{survival-package} has an implementation of ridge-penalized
estimation. On the other hand, our package \pkg{casebase} provides
penalized estimation in the context of parametric hazards. To our
knowledge, \pkg{casebase} and \pkg{rsptm2} are the only packages to
offer this functionality.

@@

Finally, several packages include functions to estimate the CIF. The
corresponding methods generally fall into two categories: transformation
of the estimated hazard function, and semi-parametric estimation of the
baseline hazard. The first category broadly corresponds to parametric
survival models, where the full hazard is explicitly modeled. Using this
estimate, the survival function and the CIF can be obtained using their
functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI}
below). Packages providing this functionality include \pkg{CFC},
\pkg{Flexsurv}, \pkg{mets}, and \pkg{survival}. Our package
\pkg{casebase} also follows this approach for both single-event and
competing-risk analyses. The second category outlined above broadly
corresponds to semi-parametric models. These models do not model the
full hazard function, and therefore the baseline hazard needs to be
estimated separately in order to estimate the survival function. This is
achieved using semi-parametric estimators (e.g.~Breslow's estimator) or
parametric estimators (e.g.~spline functions). Packages that implement
this approach include \pkg{RiskRegression}, \pkg{rstpm2}, and
\pkg{survival}. As mentioned in the introduction, a key distinguishing
factor between these two approaches is that the first category leads to
smooth estimates of the cumulative incidence function, whereas the
second category often produces estimates in the form of step-wise
functions. Providing smooth estimates of the CIF was one of the main
motivations for introducing case-base sampling in survival analysis.

@@

\begin{CodeOutput}
#> Model1 Model2 Model3 Model4 
#>    489    457    439    440
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeOutput}
#>  time     
#>     0 0.00
#>   365 0.74
\end{CodeOutput}

\begin{CodeOutput}
#> [1] 0.67
\end{CodeOutput}
\end{CodeChunk}

@@

Future work will look at some of the methodological extensions of
case-base sampling. First, to assess the quality of the model fit, we
would like to study the properties of the residuals (e.g.~Cox-Snell,
martingale). More work needs to be done to understand these residuals in
the context of the partial likelihood underlying case-base sampling. The
resulting diagnostic tools could then be integrated in this package.
Also, we are interested in extending case-base sampling to account for
interval censoring. This type of censoring is very common in
longitudinal studies, and many packages (e.g.~\pkg{SmoothHazard},
\pkg{survival} and \pkg{Rstpm2}) provide functions to account for it.
Again, we hope to include any resulting methodology as part of this
package.

@@

This report was generated on 2020-07-28 10:08:15 using the following
computational environment and dependencies:

@@

\begin{CodeOutput}
#> R version 4.0.2 (2020-06-22)
#> Platform: x86_64-pc-linux-gnu (64-bit)
#> Running under: Pop!_OS 20.04 LTS
#> 
#> Matrix products: default
#> BLAS:   /usr/lib/x86_64-linux-gnu/blas/libblas.so.3.9.0
#> LAPACK: /usr/lib/x86_64-linux-gnu/lapack/liblapack.so.3.9.0
#> 
#> locale:
#>  [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C              
#>  [3] LC_TIME=en_US.UTF-8        LC_COLLATE=en_US.UTF-8    
#>  [5] LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
#>  [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                 
#>  [9] LC_ADDRESS=C               LC_TELEPHONE=C            
#> [11] LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       
#> 
#> attached base packages:
#> [1] splines   stats     graphics  grDevices utils     datasets  methods  
#> [8] base     
#> 
#> other attached packages:
#>  [1] broom_0.7.0               dotwhisker_0.5.0         
#>  [3] timereg_1.9.6             glue_1.4.1               
#>  [5] scales_1.1.1              forcats_0.5.0            
#>  [7] stringr_1.4.0             dplyr_1.0.0              
#>  [9] purrr_0.3.4               readr_1.3.1              
#> [11] tidyr_1.1.0               ggplot2_3.3.2            
#> [13] tidyverse_1.3.0           lubridate_1.7.9          
#> [15] visreg_2.7.0              tibble_3.0.3             
#> [17] riskRegression_2020.06.25 prodlim_2019.11.13       
#> [19] flexsurv_1.1.1            survival_3.2-3           
#> [21] pracma_2.2.9              magrittr_1.5             
#> [23] kableExtra_1.1.0          glmnet_4.0-2             
#> [25] Matrix_1.2-18             data.table_1.13.0        
#> [27] cowplot_1.0.0             colorspace_1.4-1         
#> [29] casebase_0.9.0           
#> 
#> loaded via a namespace (and not attached):
#>  [1] TH.data_1.0-10      VGAM_1.1-3          ellipsis_0.3.1     
#>  [4] ggstance_0.3.4      htmlTable_2.0.1     fs_1.4.2           
#>  [7] base64enc_0.1-3     rstudioapi_0.11     farver_2.0.3       
#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-1      
#> [13] xml2_1.3.2          codetools_0.2-16    knitr_1.29         
#> [16] Formula_1.2-3       jsonlite_1.7.0      cluster_2.1.0      
#> [19] dbplyr_1.4.4        png_0.1-7           compiler_4.0.2     
#> [22] httr_1.4.2          backports_1.1.8     assertthat_0.2.1   
#> [25] cli_2.0.2           acepack_1.4.1       htmltools_0.5.0    
#> [28] quantreg_5.61       tools_4.0.2         gtable_0.3.0       
#> [31] Rcpp_1.0.5          cellranger_1.1.0    vctrs_0.3.2        
#> [34] nlme_3.1-148        conquer_1.0.1       iterators_1.0.12   
#> [37] xfun_0.16           rvest_0.3.5         lifecycle_0.2.0    
#> [40] polspline_1.1.19    muhaz_1.2.6.1       MASS_7.3-51.6      
#> [43] zoo_1.8-8           hms_0.5.3           sandwich_2.5-1     
#> [46] SparseM_1.78        RColorBrewer_1.1-2  rticles_0.14       
#> [49] yaml_2.2.1          gridExtra_2.3       rms_6.0-1          
#> [52] rpart_4.1-15        latticeExtra_0.6-29 stringi_1.4.6      
#> [55] foreach_1.5.0       checkmate_2.0.0     lava_1.6.7         
#> [58] shape_1.4.4         mets_1.2.7.1        rlang_0.4.7        
#> [61] pkgconfig_2.0.3     matrixStats_0.56.0  evaluate_0.14      
#> [64] lattice_0.20-41     labeling_0.3        htmlwidgets_1.5.1  
#> [67] cmprsk_2.2-10       tidyselect_1.1.0    deSolve_1.28       
#> [70] plyr_1.8.6          R6_2.4.1            generics_0.0.2     
#> [73] Hmisc_4.4-0         multcomp_1.4-13     DBI_1.1.0          
#> [76] withr_2.2.0         pillar_1.4.6        haven_2.3.1        
#> [79] foreign_0.8-80      mgcv_1.8-31         nnet_7.3-14        
#> [82] mstate_0.2.12       modelr_0.1.8        crayon_1.3.4       
#> [85] rmarkdown_2.3       jpeg_0.1-8.1        grid_4.0.2         
#> [88] readxl_1.3.1        blob_1.2.1          reprex_0.3.0       
#> [91] digest_0.6.25       webshot_0.5.2       numDeriv_2016.8-1.1
#> [94] stats4_4.0.2        munsell_0.5.0       viridisLite_0.3.0  
#> [97] quadprog_1.5-8
\end{CodeOutput}
\end{CodeChunk}

>>>>

Several packages implement penalized estimation for the Cox model:
\pkg{glmnet} \citeyearpar{regpathcox}, \pkg{glmpath}
\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
\pkg{riskRegression} \citeyearpar{gerds_blanche}. Moreover, some
packages also include penalized estimation in the context of Cox models
with time-varying coefficients: elastic-net penalization with
\pkg{CoxRidge} \citeyearpar{perperoglou} and \pkg{rstpm2}
\citeyearpar{clements_liu}, while \pkg{survival}
\citeyearpar{survival-package} has an implementation of ridge-penalized
estimation. On the other hand, our package \pkg{casebase} provides
penalized estimation in the context of parametric hazards. To our
knowledge, \pkg{casebase} and \pkg{rsptm2} are the only packages to
offer this functionality.

@@

Finally, several packages include functions to estimate the CIF. The
corresponding methods generally fall into two categories: transformation
of the estimated hazard function, and semi-parametric estimation of the
baseline hazard. The first category broadly corresponds to parametric
survival models, where the full hazard is explicitly modeled. Using this
estimate, the survival function and the CIF can be obtained using their
functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI}
below). Packages providing this functionality include \pkg{CFC},
\pkg{flexsurv}, \pkg{mets}, and \pkg{survival}. Our package
\pkg{casebase} also follows this approach for both single-event and
competing-risk analyses. The second category outlined above broadly
corresponds to semi-parametric models. These models do not model the
full hazard function, and therefore the baseline hazard needs to be
estimated separately in order to estimate the survival function. This is
achieved using semi-parametric estimators (e.g.~Breslow's estimator) or
parametric estimators (e.g.~spline functions). Packages that implement
this approach include \pkg{RiskRegression}, \pkg{rstpm2}, and
\pkg{survival}. As mentioned in the introduction, a key distinguishing
factor between these two approaches is that the first category leads to
smooth estimates of the cumulative incidence function, whereas the
second category often produces estimates in the form of step-wise
functions. Providing smooth estimates of the CIF was one of the main
motivations for introducing case-base sampling in survival analysis.

@@

\begin{CodeOutput}
#> Model1 Model2 Model3 Model4 
#>    491    459    444    445
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeOutput}
#>  time     
#>     0 0.00
#>   365 0.72
\end{CodeOutput}

\begin{CodeOutput}
#> [1] 0.65
\end{CodeOutput}
\end{CodeChunk}

@@

Future work will look at some of the methodological extensions of
case-base sampling. First, to assess the quality of the model fit, we
would like to study the properties of the residuals (e.g.~Cox-Snell,
martingale). More work needs to be done to understand these residuals in
the context of the partial likelihood underlying case-base sampling. The
resulting diagnostic tools could then be integrated in this package.
Also, we are interested in extending case-base sampling to account for
interval censoring. This type of censoring is very common in
longitudinal studies, and many packages (e.g.~\pkg{SmoothHazard},
\pkg{survival} and \pkg{rstpm2}) provide functions to account for it.
Again, we hope to include any resulting methodology as part of this
package.

@@

This report was generated on 2020-08-04 09:28:27 using the following
computational environment and dependencies:

@@

\begin{CodeOutput}
#> R version 4.0.2 (2020-06-22)
#> Platform: x86_64-pc-linux-gnu (64-bit)
#> Running under: Pop!_OS 20.04 LTS
#> 
#> Matrix products: default
#> BLAS:   /usr/lib/x86_64-linux-gnu/blas/libblas.so.3.9.0
#> LAPACK: /usr/lib/x86_64-linux-gnu/lapack/liblapack.so.3.9.0
#> 
#> locale:
#>  [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C              
#>  [3] LC_TIME=en_US.UTF-8        LC_COLLATE=en_US.UTF-8    
#>  [5] LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
#>  [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                 
#>  [9] LC_ADDRESS=C               LC_TELEPHONE=C            
#> [11] LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       
#> 
#> attached base packages:
#> [1] splines   stats     graphics  grDevices utils     datasets  methods  
#> [8] base     
#> 
#> other attached packages:
#>  [1] broom_0.7.0               dotwhisker_0.5.0         
#>  [3] timereg_1.9.6             glue_1.4.1               
#>  [5] scales_1.1.1              forcats_0.5.0            
#>  [7] stringr_1.4.0             dplyr_1.0.0              
#>  [9] purrr_0.3.4               readr_1.3.1              
#> [11] tidyr_1.1.0               ggplot2_3.3.2            
#> [13] tidyverse_1.3.0           lubridate_1.7.9          
#> [15] visreg_2.7.0              tibble_3.0.3             
#> [17] riskRegression_2020.06.25 prodlim_2019.11.13       
#> [19] flexsurv_1.1.1            survival_3.2-3           
#> [21] pracma_2.2.9              magrittr_1.5             
#> [23] kableExtra_1.1.0          glmnet_4.0-2             
#> [25] Matrix_1.2-18             data.table_1.13.0        
#> [27] cowplot_1.0.0             colorspace_1.4-1         
#> [29] casebase_0.9.0           
#> 
#> loaded via a namespace (and not attached):
#>  [1] TH.data_1.0-10      VGAM_1.1-3          ellipsis_0.3.1     
#>  [4] ggstance_0.3.4      htmlTable_2.0.1     fs_1.4.2           
#>  [7] base64enc_0.1-3     rstudioapi_0.11     farver_2.0.3       
#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-1      
#> [13] xml2_1.3.2          codetools_0.2-16    knitr_1.29         
#> [16] Formula_1.2-3       jsonlite_1.7.0      cluster_2.1.0      
#> [19] dbplyr_1.4.4        png_0.1-7           compiler_4.0.2     
#> [22] httr_1.4.2          backports_1.1.8     assertthat_0.2.1   
#> [25] cli_2.0.2           acepack_1.4.1       htmltools_0.5.0    
#> [28] quantreg_5.61       tools_4.0.2         gtable_0.3.0       
#> [31] Rcpp_1.0.5          cellranger_1.1.0    vctrs_0.3.2        
#> [34] nlme_3.1-148        conquer_1.0.1       iterators_1.0.12   
#> [37] xfun_0.16           rvest_0.3.5         lifecycle_0.2.0    
#> [40] polspline_1.1.19    muhaz_1.2.6.1       MASS_7.3-51.6      
#> [43] zoo_1.8-8           hms_0.5.3           sandwich_2.5-1     
#> [46] SparseM_1.78        RColorBrewer_1.1-2  rticles_0.14       
#> [49] yaml_2.2.1          gridExtra_2.3       rms_6.0-1          
#> [52] rpart_4.1-15        latticeExtra_0.6-29 stringi_1.4.6      
#> [55] foreach_1.5.0       checkmate_2.0.0     lava_1.6.7         
#> [58] shape_1.4.4         mets_1.2.7.1        rlang_0.4.7        
#> [61] pkgconfig_2.0.3     matrixStats_0.56.0  evaluate_0.14      
#> [64] lattice_0.20-41     labeling_0.3        htmlwidgets_1.5.1  
#> [67] cmprsk_2.2-10       tidyselect_1.1.0    deSolve_1.28       
#> [70] R6_2.4.1            generics_0.0.2      Hmisc_4.4-0        
#> [73] multcomp_1.4-13     DBI_1.1.0           withr_2.2.0        
#> [76] pillar_1.4.6        haven_2.3.1         foreign_0.8-80     
#> [79] mgcv_1.8-31         nnet_7.3-14         mstate_0.2.12      
#> [82] modelr_0.1.8        crayon_1.3.4        rmarkdown_2.3      
#> [85] jpeg_0.1-8.1        grid_4.0.2          readxl_1.3.1       
#> [88] blob_1.2.1          reprex_0.3.0        digest_0.6.25      
#> [91] webshot_0.5.2       numDeriv_2016.8-1.1 stats4_4.0.2       
#> [94] munsell_0.5.0       viridisLite_0.3.0   quadprog_1.5-8
\end{CodeOutput}
\end{CodeChunk}
